Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-19 Manuscript/manuscript.Rmd
manuscript edits

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational, @hong2017computational]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data that can impair and bias results.

Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to large imbalance ratios (defined as the ratio of majority to minority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple, @chen2013influence]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data.

@@

[???Now discuss sampling techniques to account for imbalanced data]

?????????????????????????????????????????????????????????????????????????[include if we decide to directly compare these methods; otherwise referenced above]
A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. Here, the authors ignored the effect of imbalanced classes as well as the use of potentially biased performance metrics.

In a similar study, Hong et al. used domain data from an hESC cell line to build a linear model and predict which genomic elements were associated with TAD boundaries [@hong2017computational]. They defined boundary segments as genomic regions before and after 150 kilobases of the boundary. To solve the imbalance problem, they randomly sampled from regions outside of the 150kb flank.  For their model, the feature space was determined by the number of counts of ChIP-seq peaks defining functional genomic elements that overlapped with the two segments. Individual models were then performed until the best model which included the concensus CTCF signal feature, followed by 10 additional elements was obtained. Here, however, it is likely that 1 iteration of random under-sampling will yield biased results, depending on how representative the under-sampled data is. Likewise, it is unclear how the performances of other balancing techniques compare (i.e. over-sampling or a combination of both under and over-sampling).
?????????????????????????????????????????????????????????????????????????

@@

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 10 kb, 25 kb, 50 kb, and 100 kb. For each resolution, we examined the inclusion of three different predictor types as the feature space for downstream modeling. We coupled each predictor type with two separate variable reduction techniques. Lastly, we performed a series of three different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the precision-recall curve (AUPRC), F1 score, and Matthew's Correlation Coefficient (MCC).

>>>>

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational; @hong2017computational; @huang2015predicting]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data that can impair and bias results.

Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several external or data level solutions. These mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset. Many such solutions have been developed in a variety of fields related to machine learning [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Likewise, it has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis,].

@@

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries including class imbalance, feature engineering, and contact map resolution. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 10 kb, 25 kb, 50 kb, and 100 kb. For each resolution, we examined the inclusion of 3 different predictor types as the feature space for downstream modeling. We then performed a series of 4 different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the precision-recall curve (AUPRC), F1 score, and Matthew's Correlation Coefficient (MCC). This process was then repeated for X additional cell lines including K562, ......
